VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms

### 2019-01-22 paper/data_template.tex
changed up some of the output to support mana's queries

Exomiser$^*$&
{% for LEVEL in LEVELS %}
    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER.TEST_RANKINGS[LEVEL])) }}&
{% endfor %}
{% for LEVEL in LEVELS %}
    {{ '%d, %d, %d' % 
        (100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER.TEST_RANKINGS[LEVEL]))
    }}
    {% if LEVEL != 'PATHOGENIC' %}
        &
    {% endif %}
{% endfor %}
\\ \hhline{=|=|=|=|=||=|=|=|=}

>>>>

Exomiser(best)&
{% for LEVEL in LEVELS %}
    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER_BEST.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER_BEST.TEST_RANKINGS[LEVEL])) }}&
{% endfor %}
{% for LEVEL in LEVELS %}
    {{ '%d, %d, %d' % 
        (100.0*np.sum(np.array(EXOMISER_BEST.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER_BEST.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER_BEST.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER_BEST.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER_BEST.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER_BEST.TEST_RANKINGS[LEVEL]))
    }}
    {% if LEVEL != 'PATHOGENIC' %}
        &
    {% endif %}
{% endfor %}
\\

Exomiser(avg)&
{% for LEVEL in LEVELS %}
    {{ '%0.1f (%0.2f)' % (np.median(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]), np.mean(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL])) }}&
{% endfor %}
{% for LEVEL in LEVELS %}
    {{ '%d, %d, %d' % 
        (100.0*np.sum(np.array(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]) <= 1) / len(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]) <= 10) / len(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]),
        100.0*np.sum(np.array(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]) <= 20) / len(EXOMISER_AVERAGE.TEST_RANKINGS[LEVEL]))
    }}
    {% if LEVEL != 'PATHOGENIC' %}
        &
    {% endif %}
{% endfor %}

\\ \hhline{=|=|=|=|=||=|=|=|=}

### 2019-01-22 paper/main.tex
changed up some of the output to support mana's queries

\title[VarSight: Prioritizing Clinically Reported Variants]{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[Holt \textit{et~al}.]{James M. Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille L. Birch\,$^{\text{\sfb 1}}$, Donna M. Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alex Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A. Wilk\,$^{\text{\sfb 1}}$, Julie A. Anderson\,$^{\text{\sfb 1}}$, Jeremy Harris\,$^{\text{\sfb 1}}$, Jacob Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A. Worthey\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA % and \\
%$^{\text{\sf 2}}$Department, Institution, City, Post Code, Country.}
}

@@

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as a ranking system and compared them to another variant prioritization algorithm and two single-measure controls. % to show how the classifiers tend to outperform the other methods.  
We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
%this was probably too detailed, need a different sentence
%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
\\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
Genome and exome sequencing are both currently being used as molecular diagnostic tools for patients with rare, undiagnosed diseases \citep{ramoni2017} (TODO: XXX: CITE MORE, CSER maybe?).  Typically, these technologies are applied clinically following the workflows consisting of blood draw, sequencing, alignment, variant calling, variant annotation, variant filtering, and variant prioritization \citep{worthey2017, roy2018}.  Then, clinical analysts usually perform the more manual processes of inspecting and then clinically reporting variants based on the known set of patient phenotypes.

In general, commonly used pipelines exist for the steps from sequencing through variant calling \citep{rehm2013, cornish2015}.  Despite differences in performance, most of these standards ingest the same information to create a list of variants from sequencing data.  In contrast, methods for variant annotation and/or variant filtering are quite diverse \citep{wang2010, hu2013, jager2014, desvignes2018}.  These methods use a wide range of input sources including but not limited to population allele frequencies \citep{lek2016}, conservation scores \citep{cooper2005, siepel2006, petrovski2013}, haploinsufficiency scores \citep{huang2010, steinberg2015}, deleteriousness scores \citep{steinberg2015, rentzsch2018}, transcript impact scores \citep{kumar2009, choi2012, adzhubei2013, dong2014, jian2014} , and previously associated disease annotation \citep{stenson2003, hamosh2005, landrum2015}.  Variant prioritization is also quite diverse with some methods relying on the variant annotations to order variants \citep{hu2013} and some relying on patient phenotype to order the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.  There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.

@@

Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.  We used the latest version of the primary clinical filter for rare disease variants to replicate the standard filtering process for patients in the UDN.  In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.  In general, the filter reduces the number of variant from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).  For the specific details on the filter used, please refer to Supplementary Documents.
%metadata available in the VarSight repository.

\subsection{Phenotype annotation}
The Codicem annotations are all agnostic of the patient phenotype.  As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases and phenotypes tied to UDN patients.  In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.  We then annotated each variant with the best scores from their corresponding gene(s).

@@

\subsection{Patient selection}
In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.  Note that secondary and/or incidental findings are specifically not included in this list.  The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic based on the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.

@@

\section{Discussion}
\subsection{Classifier Statistics}
\begin{figure*}
\centerline{
\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
}
\caption{Receiver operator curves.  This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.  On the left, we show the false positive rate against the true positive rate.  On the right, we show the recall against the precision.  Area under the receiver-operator curve (AUROC) is reported beside each method in the legend.  In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly.  LogisticRegression tends to perform slightly worse overall, but still better than either single-measure statistic.}
\label{fig:roc}
\end{figure*}

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}.  
%The values for each model are shown in Table \ref{tab:classifier_results}.  
Figure \ref{fig:roc} show receiver operator curves for FPR v. TPR and recall v. precision for the four models.  Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  All four classifiers have higher AUROC than either single-measure statistic.  However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was only 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

@@

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}

@@

For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).  We followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.  For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.  We then parsed the output JSON file created by Exomiser into a ranked order of variants.  Of note, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 20-50 variants from the pre-filtered VCF.  For primary variants that were unranked by Exomiser, we conservatively assumed that the Exomiser ranked those variants as just outside the range of what was returned (e.g. if Exomiser ranked 15 variants, we assumed all missing variants were ranked 16th).

Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.  These scores were inputs to each classifier, but also represent two common ways variants one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).  The results for the two control scores, Exomiser, and all four classifiers are shown in Table \ref{tab:performance}.

In the overall data, all four classifiers outperform the single-measure statistics and Exomiser across the board, with the one exception being Exomiser's mean (this measure is likely heavily biased due to our conservative ranking of unranked variants, harsher penalties for unranked variants would drastically change this result).  As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.  Of the classifiers tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.

\subsection{Random Forest Feature Importances}
After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.  Both random forest models calculate a feature importance array after training that sum to 1.0 (the other two classifiers do not have this information readily available).  By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.  Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ as reported by each classifier.

@@

We expect these classification algorithms could be refined in a variety of ways.  First, adding new features and/or removing unused features could lead to improvements in the algorithm.  The top two most important features were both phenotype related suggesting that a highly accurate phenotype scoring system would greatly benefit these algorithms.  Many features had a low importance, so pruning them may improve the overall results.  Additionally, some of the features represent data that is not freely available to the research community, so pruning or replacing those features with publicly accessible sources would likely influence the results.  Second, there may be a better classification algorithm for this type of data.  The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.

@@

\section*{Acknowledgements}
We are grateful for the participation of patients and family members and all of their referring clinicians. We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients. We would also like to acknowledge the UDN whole genome sequencing core headed by Dr. Shawn Levy.

@@

@@

%GERP - conservation
\bibitem[Cooper {\it et~al}., 2005]{cooper2005} Cooper, Gregory M., \textit{et al.} "Distribution and intensity of constraint in mammalian genomic sequence." {\it Genome research} 15.7 (2005): 901-913.

@@

>>>>

\title[VarSight: Prioritizing Clinically Reported Variants]{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[Holt \textit{et~al}.]{James M. Holt\,$^{\text{\sfb 1,}*}$, Brandon Wilk\,$^{\text{\sfb 1}}$, Camille L. Birch\,$^{\text{\sfb 1}}$, Donna M. Brown\,$^{\text{\sfb 1}}$, Manavalan Gajapathy\,$^{\text{\sfb 1}}$, Alexander C. Moss\,$^{\text{\sfb 1}}$, Nadiya Sosonkina\,$^{\text{\sfb 1}}$, Melissa A. Wilk\,$^{\text{\sfb 1}}$, Julie A. Anderson\,$^{\text{\sfb 1}}$, Jeremy M. Harris\,$^{\text{\sfb 1}}$, Jacob M. Kelly\,$^{\text{\sfb 1}}$, Fariba Shaterferdosian\,$^{\text{\sfb 1}}$, Angelina E. Uno-Antonison\,$^{\text{\sfb 1}}$, Arthur Weborg\,$^{\text{\sfb 1}}$, Undiagnosed Diseases Network and Elizabeth A. Worthey\,$^{\text{\sfb 1}}$}
\address{$^{\text{\sf 1}}$Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA % and \\
%$^{\text{\sf 2}}$Department, Institution, City, Post Code, Country.}
}

@@

\abstract{\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. % to show how the classifiers tend to outperform the other methods.  
We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
%this was probably too detailed, need a different sentence
%The tested classifiers had a median rank of 7.0-8.0 for all variants and 2.0-3.5 for variants reported as pathogenic.  
\\
\textbf{Availability:} The scripts used to generate results are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
Genome and exome sequencing are both currently being used as molecular diagnostic tools for patients with rare, undiagnosed diseases \citep{ramoni2017, bagnall2018, sweeney2018}.  Typically, these technologies are applied clinically following workflows consisting of blood draw, sequencing, alignment, variant calling, variant annotation, variant filtering, and variant prioritization \citep{worthey2017, roy2018}.  Then, clinical analysts usually perform the more manual processes of inspecting and then clinically reporting variants based on the known set of patient phenotypes.

In general, commonly used pipelines exist for the steps from sequencing through variant calling \citep{rehm2013, cornish2015}.  Despite differences in performance, most of these standards ingest the same information to create a list of variants from sequencing data.  In contrast, methods for variant annotation and/or variant filtering are quite diverse \citep{wang2010, hu2013, jager2014, desvignes2018}.  These methods use a wide range of input sources including but not limited to population allele frequencies \citep{lek2016}, conservation scores \citep{cooper2005, siepel2006, petrovski2013}, haploinsufficiency scores \citep{huang2010, steinberg2015}, deleteriousness scores \citep{steinberg2015, rentzsch2018}, transcript impact scores \citep{kumar2009, choi2012, adzhubei2013, dong2014, jian2014} , and previously associated disease annotation \citep{stenson2003, hamosh2005, landrum2015}.  Variant prioritization is also quite diverse with some methods relying on the variant annotations to prioritize variants \citep{hu2013} and some relying on patient phenotype to rank the variants \citep{kohler2009, yang2015, rao2018, wilk2018}.  There are also methods which combine both variant annotations and phenotype score to rank the variants \citep{singleton2014, zemojtel2014, smedley2015a}, a selection of which are benchmarked on the same simulated datasets in \cite{smedley2015b}.

@@

Codicem also performs filtering for the analysts to reduce the number of variants that are viewed through a standard clinical analysis.  We used the latest version of the primary clinical filter for rare disease variants to replicate the standard filtering process for patients in the UDN.  In short, the filter requires the following for a variant to pass through the clinical filter: sufficient total read depth, sufficient alternate read depth, low population frequency, at least one predicted effect on a transcript, at least one gene-disease association, and to not be a known, common false-positive from sequencing.  In general, the filter reduces the number of variants from the order of millions to hundreds (anecdotally, roughly 200-400 variants per proband after filtering).  For the specific details on the filter used, please refer to Supplementary Documents.
%metadata available in the VarSight repository.

\subsection{Phenotype annotation}
The Codicem annotations are all agnostic of the patient phenotype.  As noted earlier, we expect these patient phenotypes to be noisy when compared to simulated datasets due to the variety and complexity of diseases, phenotypes, and genetic heritage tied to UDN patients.  In order to incorporate patient phenotype information, we used two distinct methods to rank genes based on the Human Phenotype Ontology (HPO) \citep{kohler2018}.  We then annotated each variant with the best scores from their corresponding gene(s).

@@

\subsection{Patient selection}
In our analysis, we focused on variants that were clinically reported as ``primary", meaning the analysts believed the variant to be directly related to the patient's phenotype.  Note that secondary and/or incidental findings are specifically not included in this list.  The analysts assigned each primary variant a classification from variant of uncertain significance (VUS), likely pathogenic, or pathogenic adhering to the recommendations in the ACMG guidelines for variant classification \citep{richards2015}.

@@

\section{Results}%mana said results!
\subsection{Classifier Statistics}
\begin{figure*}
\centerline{
\includegraphics[width=0.45\textwidth]{codi_rf_roc.png}
\includegraphics[width=0.45\textwidth]{codi_rf_pr.png}
}
\caption{Receiver operator curves.  
%This figure shows the receiver operator curves for the test data for each single-measure statistic and for each trained classifier.  
This figure shows the receiver operator curves for the test data for each trained classifier.  
On the left, we show the false positive rate against the true positive rate.  On the right, we show the recall against the precision.  Area under the receiver-operator curve (AUROC) is reported beside each method in the legend.  In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly with LogisticRegression performing slightly worse overall.}
\label{fig:roc}
\end{figure*}

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}.  
%The values for each model are shown in Table \ref{tab:classifier_results}.  
Figure \ref{fig:roc} show receiver operator curves for FPR v. TPR and recall v. precision for the four models.  %Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.

From these metrics, the two random forest models and the EasyEnsembleClassifier have similar performance whereas whereas LogisticRegression performs slightly worse across the board.  
%All four classifiers have higher AUROC than either single-measure statistic.  
However, these classifier all have relatively poor performance from a precision-recall perspective (best AUROC for precision-recall was 0.2109).  This indicates that from a classification perspective, these models would identify a high number of false positives relative to the true positives unless a very conservative cutoff score was used.

@@

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}

@@

For comparison, we selected to run Exomiser \citep{smedley2015a} because it performed comparatively well with all phenotypes even in the presence of noise (see benchmarking from \citealp{smedley2015b}).  We followed the installation on their website to install Exomiser CLI v.11.0.0 along with version 1811 for hg19 data sources.  For each test case, we created a VCF file from the pre-filtered list of annotated variants created by Codicem and passed those VCFs along with the full list of patient HPO terms to Exomiser.  We then parsed the output JSON file created by Exomiser into a ranked order of variants.

Unfortunately, Exomiser did not rank every variant that was present in the filtered VCF, typically ranking only 10-50 variants from the pre-filtered VCF.  For the reported variants, Exomiser did not rank 98 of the 189 variants in the test set (58 variant of uncertain significance, 21 likely pathogenic, and 19 pathogenic).  For these unranked variants, we assigned a rank using two different methods.  The ``Exomiser(best)" method conservatively assumed that unranked primary variants were found at the next best position in the rank order (e.g. if Exomiser ranked 15 variants, we assumed all unranked variants were ranked 16th). In contrast, the ``Exomiser(avg)" method assumed that unranked primary variants were ranked at the average position for all unranked variants (e.g. if Exomiser ranked 10 of 100 variants, we assumed all unranked variants were ranked 55th).  ``Exomiser(best)" represents the best possible ordering that Exomiser could give us for unranked variants where ``Exomiser(avg)" is a more realistic expectation.

Finally, we added two control scores for comparison: CADD scaled and HPO-cosine.  These scores were inputs to each classifier, but also represent two common ways one might naively order variants after filtering (by predicted deleteriousness and by similarity to phenotype).  The results for the two control scores, both Exomiser approaches, and all four classifiers are shown in Table \ref{tab:performance}.

In the overall data, all four classifiers outperform the single-measure statistics and both Exomiser approaches across the board, with the one exception being the mean of ``Exomiser(best)" (a measure that is heavily biased by the conservative ranking of unranked variants).  As one would intuitively expect, all classifiers perform better as the returned pathogenicity increases with the EasyEnsembleClassifier ranking 36\% of pathogenic variants in the first position and 97\% of pathogenic variants in the top 20.  Of the classifiers tested, the trained EasyEnsembleClassifier performs best overall with the strongest comparative performance in the pathogenic category.

\subsection{Random Forest Feature Importances}
After training and testing each classifier, we wished to explore which features played the largest role in how the classifier functioned.  Both random forest models calculate a feature importance array that stores how important each feature is in the trained model (the other two classifiers do not have this information readily available).  The total feature importance array sums to 1.0, and higher values indicate that a feature was used more by the trained model.  By chance, we had exactly 50 parameters as input to our model, thus we would expect approximately 2\% weight to be assigned to each parameter by chance.  Table \ref{tab:features} shows each features with an average importance $\ge 0.02$ across the random forest classifiers.

@@

We expect these classification algorithms could be refined in a variety of ways.  First, adding new features and/or removing unused features could lead to improvements in the algorithm.  The top two most important features were both phenotype related suggesting that a highly accurate phenotype scoring system would greatly benefit these algorithms.  Many features had a low importance, so pruning them may improve the overall results.  Additionally, some of the features represent data that is not freely available to the research community, so pruning or replacing those features with publicly accessible sources would likely influence the results.  Second, there may be a better classification algorithm for this type of data.  The four selected classifiers were all freely available methods intended to handle the large class imbalance in the training set, but other algorithms that aren't as readily available may improve the result.  Finally, training the model on different patient populations may yield different models.  We trained on a fairly diverse patient population

@@

\section*{Acknowledgements}
We are grateful for the participation of patients and family members of the UDN and all of their referring clinicians. We would like to acknowledge all of the teams within the UDN including the coordinating center and all of the clinical sites working hard to provide definitive diagnoses for UDN patients. We would also like to acknowledge the UDN whole genome sequencing core headed by Dr. Shawn Levy.

@@

%clinical WGS example
\bibitem[Bagnall {\it et~al}., 2018]{bagnall2018} Bagnall, Richard D., \textit{et~al.} "Whole genome sequencing improves outcomes of genetic testing in patients with hypertrophic cardiomyopathy." {\it Journal of the American College of Cardiology} 72.4 (2018): 419-429.

@@

%GERP - conservation
\bibitem[Cooper {\it et~al}., 2005]{cooper2005} Cooper, Gregory M., \textit{et~al.} "Distribution and intensity of constraint in mammalian genomic sequence." {\it Genome research} 15.7 (2005): 901-913.

@@

%Example WGS in clinic
\bibitem[Sweeney {\it et~al}., 2018]{sweeney2018} Sweeney, Nathaly M., \textit{et~al.} "The case for early use of rapid whole genome sequencing in management of critically ill infants: Late diagnosis of Coffin-Siris syndrome in an infant with left congenital diaphragmatic hernia, congenital heart disease and recurrent infections." {\it Molecular Case Studies} (2018): mcs-a002469.
